Predicting Success of Motivational Interviews with Recurrent Neural Network and Probabilistic Model for Patient-Provider Communication Sequences

### 2017-08-31 paper-amia-summit-2018/amia_summit_2018.tex
fix spelling error

\section*{Introduction}
Data in the form of temporally ordered sequences of discrete or continuous observations (e.g., symbolic sequences such as notes in patient EHR, diagnostic codes, protein or DNA sequences or continuous time series, such as ECG measurements) arise in various domains of health informatics. The order of observations in a sequence is challenging to capture using features, which makes sequence classification a more challenging task than traditional classification. In general, sequence classification methods can be divided into three  categories: feature-based, distance metric-based and model-based method. Feature-based classification methods transform a sequence into a feature vector and apply a standard supervised learning algorithm, such as support vector machine \cite{leslie2004fast} or decision tree \cite{chuzhanova1998feature}. Shapelet \cite{ye2009time} and pattern \cite{kudenko1998feature, lesh1999mining} based techniques as well as hierarchical approaches \cite{nallam2016effective} have been proposed instead of standard classifiers as well. Distance-based methods measure the similarity between sequences to determine the quality of classification. The most commonly used distance function is Euclidian distance \cite{keogh2003need} with Dynamic Time Wrapping \cite{keogh2000scaling} used for more flexible matching in time series data. The third type of sequence classification methods, first create a probabilistic model of a sequence, such as Markov Chain (MC) and Hidden Markov Model \cite{rabiner1989tutorial} (HMM).

However, Markov chain cannot handle long-term dependencies in sequences. Similarly, the hidden states of HMMs follow a first-order markov chain and can select only one hidden state at each time step. Therfore, markov chain and HMMs are only capable of passing short-term dependency among sequence steps. On the other hand, Recurrent Neural Networks (RNNs) have no such limitations. RNNs are capable to capture long-term dependencies with their unlimited memory. It was shown that RNNs are very effective in sequence modeling with temporal data.  Recently, two variant of RNNs are proposed: Long Short-Term Memory\cite{graves2013speech} (LSTM) and Gated Reccurent Unit\cite{chung2014empirical} (GRU), which gave superior performance in a wide range of sequence modeling problems \cite{nion2013handwritten, lipton2015learning, choi2016doctor}.

@@

Our research addresses this problem from multiple angles. In our previous work \cite{kotov2015interpretable, hasan2016study}, we explored several machine learning methods for automatic annotation of clinical interview fragments with a specialized codebook containing a large number of patient and provider behavior codes \cite{carcone2013provider}. In this work, we propose Recurrent Neural Networks with two baseline probabilistic methods to identify patient-provider communication sequences that are likely to elicit the desired patient behavioral response (i.e. change talk or commitment language) and to dynamically estimate the likelihood of observing this desired response at any point during a clinical interview based on all coded previous patient-provider communication exchanges in the same interview. While there have been some previous qualitative studies of patient-provider dialog in a clinical setting \cite{eide2004physician}, there have been no previous work on computational modeling of annotated patient-provider communication (PPC) exchanges and predicting a desired patient behavior in a context of motivational interviews.

\section*{Methods}
\subsection*{\textit{Data collection}}
The experimental dataset for this work was constructed from the transcripts of 106 motivational interviews, which include a total of 41,764 segmented and annotated utterances. Each transcript consists of a counselor-adolescent and a counselor-caregiver session. Since the ultimate goal of a motivational interview is to increase the desire and ability of adolescents for a targeted behavior change, we considered only communication sequences from counselor-adolescent sessions and disregarded all communication sequences from counselor-caregiver sessions. The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider}, which are grouped into the adolescent and counselor code groups. Utterances were divided into successful and unsuccessful communication sequences. Successful communication sequences result in positive change talk and commitment language (a special class of change talk) statements by an adolescent, while unsuccessful sequences are the ones that result in negative change talk or commitment language and the sequences, in which no change talk or commitment language statements occur. Out of 4169 observed sequences, 3427 were positive and 742 were negative. For each of the probabilistic models (Markov chain and HMM), one model was trained using successful sequences and one model was trained using unsuccessful sequences. Statistics of experimental dataset are presented in Table~\ref{tab:data_dist} and a fragment of an adolescent session transcript is presented in Table~\ref{tab:anno_examp}. \\

@@

Our proposed baseline prediction method consists of two steps. In the first step, we model successful and unsuccessful patient-provider interactions using first- and second-order Markov Chain and Hidden Markov Model, which are popular probabilistic models for discrete observation sequences with finite vocabulary. In the second step, we classified each test sequence based on the maximum likelihood of generating that sequence from each model. Although HMM was originally developed for speech recognition \cite{rabiner1989tutorial}, it is one of the most widely used methods for sequence modeling \cite{mutsam2016maximum, eickeler1998hidden, srivastava2007hmm, won2004training, chai2001folk}. However, the latest advances in deep learning technologis shows that deep learning, in particular RNNs, provide better results than conventional machine learning methods for the task of sequence classification. Therefore, we employed two special variant of RNN models in our experiemnt: long short-term memory (LSTM) and gated recurrent unit (GRU).

\textbf {Markov Chain (MC)} is a stochastic model for randomly changing systems, which assumes that the next state of a system only depends on its current state and not on its past states (Markov property). Generally, this assumption enables reasoning and computation with the models that would otherwise be intractable. For the sequential analysis, we built two Markov models $M$ and $\overline{M}$ summarizing provider strategies and patient responses in case of successful ($M$) and unsuccessful ($\overline{M}$) motivational interviews. A Markov model $M$ can be represented as a weighted directed graph $G = (V, E, p)$, in which:
\begin{itemize}
\item $V = \{CML+, CHT+, CHT-, AMB-, BLT, LUP+, LUP-, HUP-W, ...\}$ is a set of vertices, consisting of adolescent and counselor MI behavior codes;
\item $E \subseteq V \times V$ is a set of edges corresponding to posssible transitions from one MI behavior code to the other in a sequence;
\item $p_M:E\rightarrow[0...1]$ is a function that assigns probability $p(c_i|c_j)$ to an edge between the MI behavior codes $c_i$ and $c_j$ based on maximum likelihood estimator:
\begin{equation}
P_M(c_j|c_i) = \frac{n_{c_i,c_j}}{n_{c_i}}
\end{equation}
\end{itemize}
where $n_{c_i,c_j}$ and $n_{c_i}$ are the number of times a transition between the MI behavior codes $c_i$ and $c_j$ and the code $c_i$ has been observed in the training data. Given a Markov model $M$ (such that $S\subseteq V$), the probability that a sequence of MI behavior codes $S = \{C_1,...,C_N\}$ has been generated from a Markov model $M$ is:
\begin{equation}
P_M(S) = \prod_{i=2}^N p_M(c_i|c_1,\dots,c_{i-1})=\prod_{i=2}^N p_M(c_i|c_{i-1})
\end{equation}
In the second step, we quantify the likelihood of success of a given motivational interview at a certain time point given a sequence of MI behavior codes $S$ observed prior to that point using the following formula:
\begin{equation}
p(S\rightarrow CML+) = \log\left(\frac{P_M(S)}{P_{\overline M}(S)}\right)= \sum_{i=2}^N \log p_M(c_i|c_{i-1})-\sum_{i=2}^N \log p_{\overline M}(c_i|c_{i-1})\label{eq:class}
\end{equation}
If $p(S\rightarrow CML+) > 0 $, the interview is predicted to result in positive change talk or commitment language.

@@

\textbf {Recurrent Neural Network (RNN)} is a type of Neural Network, which can send feedback from the current hidden state to the hidden state of the next time step. In this way, RNN can capture long-term dependencies for predicting the future events, which is the main advantage of RNN over feedforward Neural Network, MC, and HMM. The capability of remembering past event is very useful in motivational interviews where a behavior observed at some points in the interview is very informative for the future behaviors that will be observed. However, it was observed that RNN fails to capture long-term dependencies due to vanishing gradient problem \cite{bengio1993problem}. In order to mitigate this problem, Hochreiter et al.\cite{hochreiter1997long} proposed a special kind of RNN called Long Short Term Memory networks or simply LSTM. Later, the model was enhanced by including forget gates and the tanh activation function \cite{graves2013speech}. There are several variants of LSTM model, where a dramatic variation on the LSTM is the Gated Recurrent Unit or simply GRU \cite{cho2014properties}. GRUs are simpler than LSTM units and experimentally outperform than other models in many cases. We reiterate the mathematical formulation of GRU defined by Chung et al.\cite{chung2014empirical} as follows (Eq.~\ref{eq:firstgru}~to~\ref{eq:lastgru}):
\begin{equation}
z_t = \sigma(W_zx_t + U_zh_{t-1} + b_z)
\label{eq:firstgru}
\end{equation}
\begin{equation}
r_t = \sigma(W_rx_t + U_rh_{t-1} + b_r)
\end{equation}
\begin{equation}
\tilde h_t = tanh(W_hx_t + r_t \odot U_hh_{t-1} + b_h) 
\end{equation}
\begin{equation}
h_t = z_t \odot h_{t-1} + (1-z_t) \odot \tilde h_t
\label{eq:lastgru}
\end{equation}  
\begin{equation}
total\ loss = \alpha \cdot \frac{1}{T}\sum_{t=1}^T loss(\bar y^{(t)},y^{(t)}) + (1 - \alpha) \cdot loss(\bar y^{(T)},y^{(T)})
\label{eq:loss}
\end{equation}  
\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.90\textwidth]{figures/rnn.eps}
    \caption{\textbf{Proposed RNN model with target replication}.}
    \label{fig:rnn-model}
\end{figure}
The architecture of the proposed Recurrent Neural Network is shown in Figure~\ref{fig:rnn-model}. We used one hidden LSTM or GRU layer of 20 nodes after embedding the given myscope codes from input sequence. Then, we applied a softmax layer at each time step for predicting the target. This target replication strategy was used in several studies \cite{lipton2015learning,choi2016doctor} after introduced by Lee et al.\cite{lee2015deeply} in 2015 for convolutional neural networks. As it can be seen from equation~(\ref{eq:loss}), our loss is the combination of final loss and the mean of the losses over all sequence steps, where T is the total number of sequence step, $\bar y$ is the output at step t, and $\alpha\ \epsilon\ [0, 1]$ is a hyperparameter that determine the relative importance of intermediate outputs. LSTM and GRU are trained on 70\% of the data and validated on 20\%. The remaining 10\% of the data is used as a test set for reporting the performance of the model. Our models contain several hyperparameters such as embedding dimension, number of hidden units, learning rate, batch size, etc., which were determined by the validation set. We implemented our models in Tensorflow with adam optimizer and early stopping based on the validation loss. Experimentally, we observed that our model converges to optimal after 100 epoch and have no effect of dropout and regularization. The code for all models is publicly available at https://github.com/teanalab/myscope-sequential-analysis.

\subsection*{\textit{Evaluation metrics}}
Performance of the proposed method was evaluated in terms of precision, recall, and F-measure using 10 folds cross-validation and weighted macro-averaging of these metrics over the folds. A true positive (TP) was counted when the method correctly classified a sequence into its actual class; a false positive (FP) was counted for a class when the method incorrectly classified a sequence into that class; a false negative (FN) for an actual class of sequence was counted when the method incorrectly classified the sequence into other class. Precision of a class was defined as the ratio of the numbers of correctly classified sequences and all sequences identified as belonging to that particular class by the classifier (i.e Precision = TP / (TP + FP)). Recall of a class was defined as the ratio between the numbers of correctly classified sequences and all sequences of that particular class in the gold standard (i.e. Recall = TP / (TP + FN)). F-measure is computed as the harmonic mean of precision and recall (i.e. 2 x Precision x Recall / (Precision + Recall)). However, Accuracy is simply computed as the ratio of correctly classified sequences and total number of sequences.

>>>>

\section*{Introduction}
Data in the form of temporally ordered sequences of discrete or continuous observations (e.g., symbolic sequences such as notes in patient EHR, diagnostic codes, protein or DNA sequences or continuous time series, such as ECG measurements) arise in various domains of health informatics. The order of observations in a sequence is challenging to capture using features, which makes sequence classification a more challenging task than traditional classification. In general, sequence classification methods can be divided into three categories: feature-based, distance metric-based and model-based method. Feature-based classification methods transform a sequence into a feature vector and apply a standard supervised learning algorithm, such as support vector machine \cite{leslie2004fast} or decision tree \cite{chuzhanova1998feature}. Shapelet \cite{ye2009time} and pattern \cite{kudenko1998feature, lesh1999mining} based techniques as well as hierarchical approaches \cite{nallam2016effective} have been proposed instead of standard classifiers as well. Distance-based methods measure the similarity between sequences to determine the quality of classification. The most commonly used distance function is Euclidian distance \cite{keogh2003need} with Dynamic Time Wrapping \cite{keogh2000scaling} used for more flexible matching in time series data. The third type of sequence classification methods, first creates a probabilistic model of a sequence, such as Markov Chain (MC) and Hidden Markov Model \cite{rabiner1989tutorial} (HMM).

However, Markov chain cannot handle long-term dependencies in sequences. Similarly, the hidden states of HMMs follow a first-order markov chain and can select only one hidden state at each time step. Therefore, Markov chain and HMMs are only capable of passing short-term dependency among sequence steps. On the other hand, Recurrent Neural Networks (RNNs) have no such limitations. RNNs are capable to capture long-term dependencies with their unlimited memory. It was shown that RNNs are very effective in sequence modeling with temporal data.  Recently, two variant of RNNs is proposed: Long Short-Term Memory\cite{graves2013speech} (LSTM) and Gated Reccurent Unit\cite{chung2014empirical} (GRU), which gave superior performance in a wide range of sequence modeling problems \cite{nion2013handwritten, lipton2015learning, choi2016doctor}.

@@

Our research addresses this problem from multiple angles. In our previous work \cite{kotov2015interpretable, hasan2016study}, we explored several machine learning methods for automatic annotation of clinical interview fragments with a specialized codebook containing a large number of patient and provider behavior codes \cite{carcone2013provider}. In this work, we propose Recurrent Neural Networks with two baseline probabilistic methods to identify patient-provider communication sequences that are likely to elicit the desired patient behavioral response (i.e. change talk or commitment language) and to dynamically estimate the likelihood of observing this desired response at any point during a clinical interview based on all coded previous patient-provider communication exchanges in the same interview. While there have been some previous qualitative studies of patient-provider dialog in a clinical setting \cite{eide2004physician}, there have been no previous work on computational modeling of annotated patient-provider communication (PPC) exchanges and predicting the desired patient behavior in a context of motivational interviews.

\section*{Methods}
\subsection*{\textit{Data collection}}
The experimental dataset for this work was constructed from the transcripts of 106 motivational interviews, which include a total of 41,764 segmented and annotated utterances. Each transcript consists of a counselor-adolescent and a counselor-caregiver session. Since the ultimate goal of a motivational interview is to increase the desire and ability of adolescents for a targeted behavior change, we considered only communication sequences from counselor-adolescent sessions and disregarded all communication sequences from counselor-caregiver sessions. The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider}, which are grouped into the adolescent and counselor code groups. Utterances were divided into successful and unsuccessful communication sequences. Successful communication sequences result in positive change talk and commitment language (a special class of change talk) statements by an adolescent, while unsuccessful sequences are the ones that result in negative change talk or commitment language and the sequences, in which no change talk or commitment language statements occur. Out of 4169 observed sequences, 3427 were positive and 742 were negative. For each of the probabilistic models (Markov chain and HMM), one model was trained using successful sequences and one model was trained using unsuccessful sequences. Statistics of the experimental dataset are presented in Table~\ref{tab:data_dist} and a fragment of an adolescent session transcript is presented in Table~\ref{tab:anno_examp}. \\

@@

Our proposed baseline prediction method consists of two steps. In the first step, we model successful and unsuccessful patient-provider interactions using first- and second-order Markov Chain and Hidden Markov Model, which are popular probabilistic models for discrete observation sequences with finite vocabulary. In the second step, we classified each test sequence based on the maximum likelihood of generating that sequence from each model. Although HMM was originally developed for speech recognition \cite{rabiner1989tutorial}, it is one of the most widely used methods for sequence modeling \cite{mutsam2016maximum, eickeler1998hidden, srivastava2007hmm, won2004training, chai2001folk}. However, the latest advances in deep learning technologies show that deep learning, in particular, RNNs provide better results than conventional machine learning methods for the task of sequence classification. Therefore, we employed two special variants of RNN models in our experiemnts: long short-term memory (LSTM) and gated recurrent unit (GRU).

\textbf {Markov Chain (MC)} is a stochastic model for randomly changing systems, which assumes that the next state of a system only depends on its current state and not on its past states (Markov property). Generally, this assumption enables reasoning and computation with the models that would otherwise be intractable. For the sequential analysis, we built two Markov models $M$ and $\overline{M}$ summarizing provider strategies and patient responses in case of successful ($M$) and unsuccessful ($\overline{M}$) motivational interviews. A Markov model $M$ can be represented as a weighted directed graph $G = (V, E, p)$, in which:
\begin{itemize}
\item $V = \{CML+, CHT+, CHT-, AMB-, BLT, LUP+, LUP-, HUP-W, ...\}$ is a set of vertices, consisting of adolescent and counselor MI behavior codes;
\item $E \subseteq V \times V$ is a set of edges corresponding to possible transitions from one MI behavior code to the other in a sequence;
\item $p_M:E\rightarrow[0...1]$ is a function that assigns probability $p(c_i|c_j)$ to an edge between the MI behavior codes $c_i$ and $c_j$ based on maximum likelihood estimator:
\begin{equation}
P_M(c_j|c_i) = \frac{n_{c_i,c_j}}{n_{c_i}}
\end{equation}
\end{itemize}
where $n_{c_i,c_j}$ and $n_{c_i}$ are the number of times a transition between the MI behavior codes $c_i$ and $c_j$ and the code $c_i$ has been observed in the training data. Given a Markov model $M$ (such that $S\subseteq V$), the probability that a sequence of MI behavior codes $S = \{C_1,...,C_N\}$ has been generated from a Markov model $M$ is:
\begin{equation}
P_M(S) = \prod_{i=2}^N p_M(c_i|c_1,\dots,c_{i-1})=\prod_{i=2}^N p_M(c_i|c_{i-1})
\end{equation}
In the second step, we quantify the likelihood of success of a given motivational interview at a certain time point given a sequence of MI behavior codes $S$ observed prior to that point using the following formula:
\begin{equation}
p(S\rightarrow CML+) = \log\left(\frac{P_M(S)}{P_{\overline M}(S)}\right)= \sum_{i=2}^N \log p_M(c_i|c_{i-1})-\sum_{i=2}^N \log p_{\overline M}(c_i|c_{i-1})\label{eq:class}
\end{equation}
If $p(S\rightarrow CML+) > 0 $, the interview is predicted to result in positive change talk or commitment language.

@@

\textbf {Recurrent Neural Network (RNN)} is a type of Neural Network, which can send feedback from the current hidden state to the hidden state of the next time step. In this way, RNN can capture long-term dependencies for predicting the future events, which is the main advantage of RNN over feedforward Neural Network, MC, and HMM. The capability of remembering past event is very useful in motivational interviews where a behavior observed at some points in the interview is very informative for the future behaviors that will be observed. However, it was observed that RNN fails to capture long-term dependencies due to vanishing gradient problem \cite{bengio1993problem}. In order to mitigate this problem, Hochreiter et al.\cite{hochreiter1997long} proposed a special kind of RNN called Long Short Term Memory networks or simply LSTM. Later, the model was enhanced by including forget gates and the tanh activation function \cite{graves2013speech}. There are several variants of LSTM model, where a dramatic variation on the LSTM is the Gated Recurrent Unit or simply GRU \cite{cho2014properties}. GRUs are simpler than LSTM units and experimentally outperform than other models in many cases. We reiterate the mathematical formulation of GRU defined by Chung et al.\cite{chung2014empirical} as follows (Eq.~\ref{eq:firstgru}~to~\ref{eq:lastgru}):
\begin{equation}
z_t = \sigma(W_zx_t + U_zh_{t-1} + b_z)
\label{eq:firstgru}
\end{equation}
\begin{equation}
r_t = \sigma(W_rx_t + U_rh_{t-1} + b_r)
\end{equation}
\begin{equation}
\tilde h_t = tanh(W_hx_t + r_t \odot U_hh_{t-1} + b_h) 
\end{equation}
\begin{equation}
h_t = z_t \odot h_{t-1} + (1-z_t) \odot \tilde h_t
\label{eq:lastgru}
\end{equation}  
\begin{equation}
total\ loss = \alpha \cdot \frac{1}{T}\sum_{t=1}^T loss(\bar y^{(t)},y^{(t)}) + (1 - \alpha) \cdot loss(\bar y^{(T)},y^{(T)})
\label{eq:loss}
\end{equation}  
\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.90\textwidth]{figures/rnn.eps}
    \caption{\textbf{Proposed RNN model with target replication}.}
    \label{fig:rnn-model}
\end{figure}
The architecture of the proposed Recurrent Neural Network is shown in Figure~\ref{fig:rnn-model}. We used one hidden LSTM or GRU layer of 20 nodes after embedding the given myscope codes from the input sequence. Then, we applied a softmax layer at each time step for predicting the target. This target replication strategy was used in several studies \cite{lipton2015learning,choi2016doctor} after introduced by Lee et al.\cite{lee2015deeply} in 2015 for convolutional neural networks. As it can be seen from equation~(\ref{eq:loss}), our loss is the combination of final loss and the mean of the losses over all sequence steps, where T is the total number of sequence step, $\bar y$ is the output at step t, and $\alpha\ \epsilon\ [0, 1]$ is a hyperparameter that determines the relative importance of intermediate outputs. LSTM and GRU are trained on 70\% of the data and validated on 20\%. The remaining 10\% of the data is used as a test set for reporting the performance of the model. Our models contain several hyperparameters such as embedding dimension, number of hidden units, learning rate, batch size, etc., which were determined by the validation set. We implemented our models in Tensorflow with adam optimizer and early stopping based on the validation loss. Experimentally, we observed that our model converges to optimal after 100 epoch and have no effect of dropout and regularization. The code for all models is publicly available at https://github.com/teanalab/myscope-sequential-analysis.

\subsection*{\textit{Evaluation metrics}}
Performance of the proposed method was evaluated in terms of precision, recall, and F-measure using 10 folds cross-validation and weighted macro-averaging of these metrics over the folds. A true positive (TP) was counted when the method correctly classified a sequence into its actual class; a false positive (FP) was counted for a class when the method incorrectly classified a sequence into that class; a false negative (FN) for an actual class of sequence was counted when the method incorrectly classified the sequence into other class. The Precision of a class was defined as the ratio of the numbers of correctly classified sequences and all sequences identified as belonging to that particular class by the classifier (i.e Precision = TP / (TP + FP)). Recall of a class was defined as the ratio between the numbers of correctly classified sequences and all sequences of that particular class in the gold standard (i.e. Recall = TP / (TP + FN)). F-measure is computed as the harmonic mean of precision and recall (i.e. 2 x Precision x Recall / (Precision + Recall)). However, Accuracy is simply computed as the ratio of correctly classified sequences and a total number of sequences.
